Event History Analysis for psychological time-to-event data: A tutorial in R with examples in Bayesian and frequentist workflows

### 2024-11-26 manuscript/manuscript.tex
update Tutorial 2a and ms.
update ms.

\caption{Discrete-time hazard and survivor functions. Discrete time-to-event data were simulated for 200 trials of 1 experimental condition. While the hazard function is the vehicle for inferring the time course of cognitive processes, the survival probability S(t-1) can help to qualify or provide context to the interpretation of the hazard probability h(t). For example, the high hazard of .60 = h(t=6) is experienced only by .44 = S(t-1=5) percent of the trials.
Because the survivor function is a decreasing function of time, the error bars in later parts of the hazard function will always be wider and less precise compared to earlier parts.}\label{fig:plot2}
\end{figure}

@@

\begin{itemize}
\tightlist
\item
  1a. Wrangle raw data and calculate descriptive stats for one independent variable.
\item
  1b. Wrangle raw data and calculate descriptive stats for two independent variables.
\item
  2a. Bayesian multilevel modeling for h(t)
\item
  2b. Bayesian multilevel modeling for ca(t)
\item
  3a. Frequentist multilevel modeling for h(t)
\item
  3b. Frequentist multilevel modeling for ca(t)
\end{itemize}

Plannng (T4) - if we get a simulation and power analysis script working, which we are happy with then we could include it here.

@@

{\centering \includegraphics[width=12in,height=0.67\textheight,]{../Tutorial_1_descriptive_stats/figures/Plot_for_subject6_PanisSchmidt}

@@

To this end, Tutorial 1b illustrates how to calculate and plot the descriptive statistics for the full data set of Experiment 1 of Panis and Schmidt (2016), which includes two independent variables: mask type and prime type. As we use the same functional programming approach as in Tutorial 1a, we simply present the sample-based functions for participant 6 as part of Tutorial 1b for those that are interested.

\subsection{4.3 Tutorial 2a: Fitting Bayesian hazard models to time-to-event data}\label{tutorial-2a-fitting-bayesian-hazard-models-to-time-to-event-data}

@@

There are several analytic decisions one has to make when fitting a hazard model. First, one has to select an analysis time window, i.e., a contiguous set of bins for which there is enough data for each participant. Second, given that the dependent variable (event occurrence) is binary, one has to select a link function (see part C in the supplementary material). The cloglog link is preferred over the logit link when events can occur in principle at any time point within a bin, which is the case for RT data (Singer \& Willett, 2003). Third, one has to choose a specification of the effect of discrete TIME (i.e., the time bin index t) in a selected baseline condition. One can choose a general specification (one intercept per bin) or a functional specification, such as a polynomial one (compare model 1 with models 2, 3, and 4 below; see also part D of the supplementary material).

@@

In this tutorial we will fit four Bayesian multilevel models (i.e., generalized linear mixed models) that differ in complexity to the person-trial-bin oriented data set that we created in Tutorial 1a. We decided to select the analysis time window (200,600{]} and the cloglog link. The data is prepared as follows.

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# load person{-}trial{-}bin oriented data set}
\NormalTok{ptb\_data }\OtherTok{\textless{}{-}} \FunctionTok{read\_csv}\NormalTok{(}\StringTok{"../Tutorial\_1\_descriptive\_stats/data/inputfile\_hazard\_modeling.csv"}\NormalTok{)}
\CommentTok{\# select analysis time window: (200,600] with 10 bins (time bin ranks 6 to 15)}
\NormalTok{ptb\_data }\OtherTok{\textless{}{-}}\NormalTok{ ptb\_data }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(period }\SpecialCharTok{\textgreater{}} \DecValTok{5}\NormalTok{)}
\CommentTok{\# create factor condition, with "blank" as the reference level}
\NormalTok{ptb\_data }\OtherTok{\textless{}{-}}\NormalTok{ ptb\_data }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{mutate}\NormalTok{(}\AttributeTok{condition =} \FunctionTok{factor}\NormalTok{(condition, }\AttributeTok{labels =} \FunctionTok{c}\NormalTok{(}\StringTok{"blank"}\NormalTok{, }\StringTok{"congruent"}\NormalTok{,}\StringTok{"incongruent"}\NormalTok{)))}
\CommentTok{\# center TIME (variable period) on bin 9, and variable trial on number 1000 and rescale; Add dummy variables for each bin.}
\NormalTok{ptb\_data }\OtherTok{\textless{}{-}}\NormalTok{ ptb\_data }\SpecialCharTok{\%\textgreater{}\%} 
        \FunctionTok{mutate}\NormalTok{(}\AttributeTok{period\_9 =}\NormalTok{ period }\SpecialCharTok{{-}} \DecValTok{9}\NormalTok{,}
               \AttributeTok{trial\_c =}\NormalTok{ (trial }\SpecialCharTok{{-}} \DecValTok{1000}\NormalTok{)}\SpecialCharTok{/}\DecValTok{1000}\NormalTok{,}
               \AttributeTok{d6  =} \FunctionTok{if\_else}\NormalTok{(period }\SpecialCharTok{==} \DecValTok{6}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{),}
               \AttributeTok{d7  =} \FunctionTok{if\_else}\NormalTok{(period }\SpecialCharTok{==} \DecValTok{7}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{),}
               \AttributeTok{d8  =} \FunctionTok{if\_else}\NormalTok{(period }\SpecialCharTok{==} \DecValTok{8}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{),}
               \AttributeTok{d9  =} \FunctionTok{if\_else}\NormalTok{(period }\SpecialCharTok{==} \DecValTok{9}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{),}
               \AttributeTok{d10 =} \FunctionTok{if\_else}\NormalTok{(period }\SpecialCharTok{==} \DecValTok{10}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{),}
               \AttributeTok{d11 =} \FunctionTok{if\_else}\NormalTok{(period }\SpecialCharTok{==} \DecValTok{11}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{),}
               \AttributeTok{d12 =} \FunctionTok{if\_else}\NormalTok{(period }\SpecialCharTok{==} \DecValTok{12}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{),}
               \AttributeTok{d13 =} \FunctionTok{if\_else}\NormalTok{(period }\SpecialCharTok{==} \DecValTok{13}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{),}
               \AttributeTok{d14 =} \FunctionTok{if\_else}\NormalTok{(period }\SpecialCharTok{==} \DecValTok{14}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{),}
               \AttributeTok{d15 =} \FunctionTok{if\_else}\NormalTok{(period }\SpecialCharTok{==} \DecValTok{15}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

To get the posterior distribution of each model parameter given the data, we need to specify a prior distribution for each parameter. The middle column of Figure 12 in part E of the supplementary material shows seven examples of prior distributions on the logit and/or cloglog scales.

While a normal distribution with relatively large variance is often used as a weakly informative prior for continuous dependent variables, rows A and B in Figure 12 show that specifying such distributions on the logit and cloglog scales leads to rather informative distributions on the original probability scale, as most mass is pushed to probabilities of 0 and 1. The other rows in Figure 12 show prior distributions on the logit and cloglog scale that we use instead.

\subsubsection{4.3.3 Model 1: A general specification of TIME, and main effects of congruency and trial number}\label{model-1-a-general-specification-of-time-and-main-effects-of-congruency-and-trial-number}

When you do not want to make assumptions about the shape of the hazard function in the selected baseline condition, or its shape is not smooth but irregular, then you can use a general specification of TIME, i.e., fit one intercept per time bin. In this first model, we use a general specification of TIME for the selected baseline condition (blank prime), and assume that the effects of prime-target congruency and trial number are proportional and additive, and that the effect of trial number is linear.
Before we fit model 1, we remove unnecessary columns from the data, and specify our priors. In the code of Tutorial 2a, model M1 is specified as follows.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{model\_M1 }\OtherTok{\textless{}{-}}
   \FunctionTok{brm}\NormalTok{(}\AttributeTok{data =}\NormalTok{ M1\_data,}
       \AttributeTok{family =} \FunctionTok{binomial}\NormalTok{(}\AttributeTok{link=}\StringTok{"cloglog"}\NormalTok{),}
\NormalTok{       event }\SpecialCharTok{|} \FunctionTok{trials}\NormalTok{(}\DecValTok{1}\NormalTok{) }\SpecialCharTok{\textasciitilde{}} \DecValTok{0} \SpecialCharTok{+}\NormalTok{ d6 }\SpecialCharTok{+}\NormalTok{ d7 }\SpecialCharTok{+}\NormalTok{ d8 }\SpecialCharTok{+}\NormalTok{ Intercept }\SpecialCharTok{+}\NormalTok{ d10 }\SpecialCharTok{+}\NormalTok{ d11 }\SpecialCharTok{+}\NormalTok{ d12 }\SpecialCharTok{+}\NormalTok{ d13 }\SpecialCharTok{+}\NormalTok{ d14 }\SpecialCharTok{+}\NormalTok{ d15 }\SpecialCharTok{+} 
\NormalTok{                           condition }\SpecialCharTok{+}\NormalTok{ trial\_c }\SpecialCharTok{+}
\NormalTok{                           (d6 }\SpecialCharTok{+}\NormalTok{ d7 }\SpecialCharTok{+}\NormalTok{ d8 }\SpecialCharTok{+} \DecValTok{1} \SpecialCharTok{+}\NormalTok{ d10 }\SpecialCharTok{+}\NormalTok{ d11 }\SpecialCharTok{+}\NormalTok{ d12 }\SpecialCharTok{+}\NormalTok{ d13 }\SpecialCharTok{+}\NormalTok{ d14 }\SpecialCharTok{+}\NormalTok{ d15 }\SpecialCharTok{+} 
\NormalTok{                           condition }\SpecialCharTok{+}\NormalTok{ trial\_c }\SpecialCharTok{|}\NormalTok{ pid),}
       \AttributeTok{prior =}\NormalTok{ priors\_M1,}
       \AttributeTok{chains =} \DecValTok{4}\NormalTok{, }\AttributeTok{cores =} \DecValTok{4}\NormalTok{, }\AttributeTok{iter =} \DecValTok{3000}\NormalTok{, }\AttributeTok{warmup =} \DecValTok{1000}\NormalTok{,}
       \AttributeTok{control =} \FunctionTok{list}\NormalTok{(}\AttributeTok{adapt\_delta =} \FloatTok{0.999}\NormalTok{, }\AttributeTok{step\_size =} \FloatTok{0.04}\NormalTok{, }\AttributeTok{max\_treedepth =} \DecValTok{12}\NormalTok{),}
       \AttributeTok{seed =} \DecValTok{12}\NormalTok{, }\AttributeTok{init =} \StringTok{"0"}\NormalTok{,}
       \AttributeTok{file =} \StringTok{"Tutorial\_2\_Bayesian/models/model\_M1"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

After selecting the binomial family and the cloglog link, the model formula is specified. The fixed effects include 9 dummy variables, the explicit Intercept variable (which represents bin 9 in this example), and the main effects of prime-target congruency (variable condition) and centered trial number (variable trial\_c). Each of these effects is allowed to vary across individuals (variable pid).
Estimating model M1 took about 70 minutes on a MacBook Pro (Sonoma 14.6.1 OS, 18GB Memory, M3 Pro Chip).

\subsubsection{4.3.4 Model 2: A polynomial specification of TIME, and main effects of congruency and trial number}\label{model-2-a-polynomial-specification-of-time-and-main-effects-of-congruency-and-trial-number}

When the shape of the hazard function is rather smooth, as it is for behavioral RT data, one can fit a more parsimonious model by using a polynomial specification of TIME. For our second example model, we thus use a third-order polynomial specification of TIME for the selected baseline condition (blank prime), and again assume that the effects of prime-target congruency and centered trial number are proportional and additive, and that the effect of trial number is linear. The model formula for model M2 looks as follows.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{event }\SpecialCharTok{|} \FunctionTok{trials}\NormalTok{(}\DecValTok{1}\NormalTok{) }\SpecialCharTok{\textasciitilde{}} \DecValTok{0} \SpecialCharTok{+}\NormalTok{ Intercept }\SpecialCharTok{+}\NormalTok{ period\_9 }\SpecialCharTok{+} \FunctionTok{I}\NormalTok{(period\_9}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{) }\SpecialCharTok{+} \FunctionTok{I}\NormalTok{(period\_9}\SpecialCharTok{\^{}}\DecValTok{3}\NormalTok{) }\SpecialCharTok{+}
\NormalTok{                           condition }\SpecialCharTok{+}\NormalTok{ trial\_c }\SpecialCharTok{+}
\NormalTok{                           (}\DecValTok{1} \SpecialCharTok{+}\NormalTok{ period\_9 }\SpecialCharTok{+} \FunctionTok{I}\NormalTok{(period\_9}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{) }\SpecialCharTok{+} \FunctionTok{I}\NormalTok{(period\_9}\SpecialCharTok{\^{}}\DecValTok{3}\NormalTok{) }\SpecialCharTok{+}
\NormalTok{                           condition }\SpecialCharTok{+}\NormalTok{ trial\_c }\SpecialCharTok{|}\NormalTok{ pid),}
\end{Highlighting}
\end{Shaded}

Because TIME is centered on bin 9, and trial number on trial 1000, the Intercept represents the cloglog-hazard in bin 9 for the blank prime condition in trial 1000. Estimating model M2 took about 2.5 hours.

\subsubsection{4.3.5 Model 3: A polynomial specification of TIME, and relaxing the proportionality assumption}\label{model-3-a-polynomial-specification-of-time-and-relaxing-the-proportionality-assumption}

So far, we assumed that the effect of our predictors prime-target congruecy and centered trial number are the same in each time bin. However, the descriptive plots (e.g., Figure 4) suggest that the effect of prime-target congruency varies across time bins. Previous research has shown that psychological effects typically change over time (Panis, 2020; Panis, Moran, et al., 2020; Panis \& Schmidt, 2022; Panis et al., 2017; Panis \& Wagemans, 2009).
For the third model, we thus use a third-order polynomial specification of TIME for the baseline condition (blank prime), and relax the proportionality assumption for the predictor variables prime-target congruency (variable condition) and centered trial number (variable trial\_c).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{        event }\SpecialCharTok{|} \FunctionTok{trials}\NormalTok{(}\DecValTok{1}\NormalTok{) }\SpecialCharTok{\textasciitilde{}} \DecValTok{0} \SpecialCharTok{+}\NormalTok{ Intercept }\SpecialCharTok{+} 
\NormalTok{                            condition}\SpecialCharTok{*}\NormalTok{period\_9 }\SpecialCharTok{+}  
\NormalTok{                            condition}\SpecialCharTok{*}\FunctionTok{I}\NormalTok{(period\_9}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{) }\SpecialCharTok{+}  
\NormalTok{                            condition}\SpecialCharTok{*}\FunctionTok{I}\NormalTok{(period\_9}\SpecialCharTok{\^{}}\DecValTok{3}\NormalTok{) }\SpecialCharTok{+} 
\NormalTok{                            trial\_c}\SpecialCharTok{*}\NormalTok{period\_9 }\SpecialCharTok{+}  
\NormalTok{                            trial\_c}\SpecialCharTok{*}\FunctionTok{I}\NormalTok{(period\_9}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{) }\SpecialCharTok{+}  
\NormalTok{                            trial\_c}\SpecialCharTok{*}\FunctionTok{I}\NormalTok{(period\_9}\SpecialCharTok{\^{}}\DecValTok{3}\NormalTok{) }\SpecialCharTok{+} 
\NormalTok{                            (}\DecValTok{1} \SpecialCharTok{+}\NormalTok{ condition}\SpecialCharTok{*}\NormalTok{period\_9 }\SpecialCharTok{+} 
\NormalTok{                            condition}\SpecialCharTok{*}\FunctionTok{I}\NormalTok{(period\_9}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{) }\SpecialCharTok{+} 
\NormalTok{                            condition}\SpecialCharTok{*}\FunctionTok{I}\NormalTok{(period\_9}\SpecialCharTok{\^{}}\DecValTok{3}\NormalTok{) }\SpecialCharTok{+} 
\NormalTok{                            trial\_c}\SpecialCharTok{*}\NormalTok{period\_9 }\SpecialCharTok{+}  
\NormalTok{                            trial\_c}\SpecialCharTok{*}\FunctionTok{I}\NormalTok{(period\_9}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{) }\SpecialCharTok{+}  
\NormalTok{                            trial\_c}\SpecialCharTok{*}\FunctionTok{I}\NormalTok{(period\_9}\SpecialCharTok{\^{}}\DecValTok{3}\NormalTok{) }\SpecialCharTok{|}\NormalTok{ pid), }
\end{Highlighting}
\end{Shaded}

Note that duplicate terms across the interaction terms in the model formula (e.g., condition) are ignored. Estimating model M3 took about 4.5 hours.

\subsubsection{4.3.6 Model 4: A polynomial specification of TIME, and relaxing all three assumptions}\label{model-4-a-polynomial-specification-of-time-and-relaxing-all-three-assumptions}

Based on previous work (e.g., Panis, 2020) we expect nonlinear effects of trial number on the hazard of response occurrence. We thus relax all three assumptions in model 4. We add a squared term for the continuous predictor centered trial number -- I(trial\_c\^{}2) -- and include interaction terms. For example, how the effect of congruent primes changes across time bins within a trial might change across the trials within an experiment.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{event }\SpecialCharTok{|} \FunctionTok{trials}\NormalTok{(}\DecValTok{1}\NormalTok{) }\SpecialCharTok{\textasciitilde{}} \DecValTok{0} \SpecialCharTok{+}\NormalTok{ Intercept }\SpecialCharTok{+} 
\NormalTok{                           condition}\SpecialCharTok{*}\NormalTok{period\_9}\SpecialCharTok{*}\NormalTok{trial\_c }\SpecialCharTok{+}  
\NormalTok{                           condition}\SpecialCharTok{*}\NormalTok{period\_9}\SpecialCharTok{*}\FunctionTok{I}\NormalTok{(trial\_c}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{) }\SpecialCharTok{+}
\NormalTok{                           condition}\SpecialCharTok{*}\FunctionTok{I}\NormalTok{(period\_9}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}\SpecialCharTok{*}\NormalTok{trial\_c }\SpecialCharTok{+} 
\NormalTok{                           condition}\SpecialCharTok{*}\FunctionTok{I}\NormalTok{(period\_9}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}\SpecialCharTok{*}\FunctionTok{I}\NormalTok{(trial\_c}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{) }\SpecialCharTok{+} 
\NormalTok{                           condition}\SpecialCharTok{*}\FunctionTok{I}\NormalTok{(period\_9}\SpecialCharTok{\^{}}\DecValTok{3}\NormalTok{) }\SpecialCharTok{+} 
\NormalTok{                           trial\_c}\SpecialCharTok{*}\FunctionTok{I}\NormalTok{(period\_9}\SpecialCharTok{\^{}}\DecValTok{3}\NormalTok{) }\SpecialCharTok{+} 
\NormalTok{                           (}\DecValTok{1} \SpecialCharTok{+}\NormalTok{  condition}\SpecialCharTok{*}\NormalTok{period\_9}\SpecialCharTok{*}\NormalTok{trial\_c }\SpecialCharTok{+} 
\NormalTok{                                 condition}\SpecialCharTok{*}\NormalTok{period\_9}\SpecialCharTok{*}\FunctionTok{I}\NormalTok{(trial\_c}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{) }\SpecialCharTok{+}  
\NormalTok{                                 condition}\SpecialCharTok{*}\FunctionTok{I}\NormalTok{(period\_9}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}\SpecialCharTok{*}\NormalTok{trial\_c }\SpecialCharTok{+} 
\NormalTok{                                 condition}\SpecialCharTok{*}\FunctionTok{I}\NormalTok{(period\_9}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}\SpecialCharTok{*}\FunctionTok{I}\NormalTok{(trial\_c}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)   }
\NormalTok{                                 condition}\SpecialCharTok{*}\FunctionTok{I}\NormalTok{(period\_9}\SpecialCharTok{\^{}}\DecValTok{3}\NormalTok{) }\SpecialCharTok{+} 
\NormalTok{                                 trial\_c}\SpecialCharTok{*}\FunctionTok{I}\NormalTok{(period\_9}\SpecialCharTok{\^{}}\DecValTok{3}\NormalTok{) }\SpecialCharTok{|}\NormalTok{ pid)}
\end{Highlighting}
\end{Shaded}

Again, duplicate terms across the interaction terms in the model formula are ignored. Estimating model M4 took about 8 hours.

>>>>

\caption{Discrete-time hazard and survivor functions. Discrete time-to-event data were simulated for 200 trials of 1 experimental condition. While the hazard function is the vehicle for inferring the time course of cognitive processes, the survival probability S(t-1) can help to qualify or provide context to the interpretation of the hazard probability h(t). For example, the high hazard of .60 = h(t=6) is experienced only by 44 percent of the trials, as S(t-1=5) = .44.
Because the survivor function is a decreasing function of time, the error bars in later parts of the hazard function will always be wider and less precise compared to earlier parts.}\label{fig:plot2}
\end{figure}

@@

\begin{itemize}
\tightlist
\item
  1a. Wrangle raw data and calculate descriptive stats for one independent variable
\item
  1b. Wrangle raw data and calculate descriptive stats for two independent variables
\item
  2a. Bayesian multilevel modeling for h(t)
\item
  2b. Bayesian multilevel modeling for ca(t)
\item
  3a. Frequentist multilevel modeling for h(t)
\item
  3b. Frequentist multilevel modeling for ca(t)
\item
  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \setcounter{enumi}{3}
  \tightlist
  \item
    Simulation and power analysis for planning experiments
  \end{enumerate}
\end{itemize}

@@

{\centering \includegraphics[width=21.33in,height=0.67\textheight,]{../Tutorial_1_descriptive_stats/figures/Plot_for_subject6_PanisSchmidt}

@@

To this end, Tutorial 1b illustrates how to calculate and plot the descriptive statistics for the full data set of Experiment 1 of Panis and Schmidt (2016), which includes two independent variables: mask type and prime type. As we use the same functional programming approach as in Tutorial 1a, we simply present the sample-based functions for each participant as part of Tutorial 1b for those that are interested.

\subsection{4.3 Tutorial 2a: Fitting Bayesian hazard models to discrete time-to-event data}\label{tutorial-2a-fitting-bayesian-hazard-models-to-discrete-time-to-event-data}

@@

There are several analytic decisions one has to make when fitting a discrete-time hazard model. First, one has to select an analysis time window, i.e., a contiguous set of bins for which there is enough data for each participant. Second, given that the dependent variable (event occurrence) is binary, one has to select a link function (see part C in the supplementary material). The cloglog link is preferred over the logit link when events can occur in principle at any time point within a bin, which is the case for RT data (Singer \& Willett, 2003). Third, one has to choose whether to treat discrete TIME (i.e., the time bin index t) as a discrete or continuous predictor. And when you treat time as a discrete predictor, you can choose between reference coding and index coding.

@@

In tutorial\_2a.Rmd we fit several Bayesian multilevel models (i.e., generalized linear mixed models) that differ in complexity to the person-trial-bin oriented data set that we created in Tutorial 1a. We decided to select the analysis time window (200,600{]} and the cloglog link. Below, we shortly discuss three of these models. The person-trial-bin data set is prepared as follows.

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# read in the file we saved in tutorial 1a}
\NormalTok{ptb\_data }\OtherTok{\textless{}{-}} \FunctionTok{read\_csv}\NormalTok{(}\StringTok{"Tutorial\_1\_descriptive\_stats/data/inputfile\_hazard\_modeling.csv"}\NormalTok{)}

\NormalTok{ptb\_data }\OtherTok{\textless{}{-}}\NormalTok{ ptb\_data }\SpecialCharTok{\%\textgreater{}\%} 
\CommentTok{\# select analysis time range: (200,600] with 10 bins (time bin ranks 6 to 15)}
\FunctionTok{filter}\NormalTok{(period }\SpecialCharTok{\textgreater{}} \DecValTok{5}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
       \CommentTok{\# continuous predictor for TIME named "period\_9", centered on bin 9}
\FunctionTok{mutate}\NormalTok{(}\AttributeTok{period\_9 =}\NormalTok{ period }\SpecialCharTok{{-}} \DecValTok{9}\NormalTok{,}
       \CommentTok{\# categorical predictor for TIME named "timebin" with index coding}
       \AttributeTok{timebin =} \FunctionTok{factor}\NormalTok{(period,}\AttributeTok{levels=}\FunctionTok{c}\NormalTok{(}\DecValTok{6}\SpecialCharTok{:}\DecValTok{15}\NormalTok{)),}
       \CommentTok{\# factor "condition" using reference coding, with "blank" as the reference level}
       \AttributeTok{condition =} \FunctionTok{factor}\NormalTok{(condition, }\AttributeTok{labels =} \FunctionTok{c}\NormalTok{(}\StringTok{"blank"}\NormalTok{, }\StringTok{"congruent"}\NormalTok{,}\StringTok{"incongruent"}\NormalTok{)),}
       \CommentTok{\# categorical predictor "prime" with index coding}
       \AttributeTok{prime =} \FunctionTok{ifelse}\NormalTok{(condition}\SpecialCharTok{==}\StringTok{"blank"}\NormalTok{,}\DecValTok{1}\NormalTok{, }\FunctionTok{ifelse}\NormalTok{(condition}\SpecialCharTok{==}\StringTok{"congruent"}\NormalTok{,}\DecValTok{2}\NormalTok{,}\DecValTok{3}\NormalTok{)),}
       \AttributeTok{prime =} \FunctionTok{factor}\NormalTok{(prime,}\AttributeTok{levels=}\FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{,}\DecValTok{3}\NormalTok{)))}
\end{Highlighting}
\end{Shaded}

@@

To get the posterior distribution of each model parameter given the data, we need to specify prior distributions for the model parameters which reflect our prior beliefs.
In Tutorial\_2a.Rmd we perform some prior predictive checks to make sure our selected prior distributions reflect our prior beliefs.

The middle column of Figure 12 in part E of the supplementary material shows seven examples of prior distributions for an intercept on the logit and/or cloglog scales. While a normal distribution with relatively large variance is often used as a weakly informative prior for continuous dependent variables, rows A and B in Figure 12 show that specifying such distributions on the logit and cloglog scales actually leads to rather informative distributions on the original probability scale, as most mass is pushed to probabilities of 0 and 1.

\subsubsection{4.3.3 Model M0i: A general specification of TIME}\label{model-m0i-a-general-specification-of-time}

When you do not want to make assumptions about the shape of the hazard function, or its shape is not smooth but irregular, then you can use a general specification of TIME, i.e., fit one grand intercept per time bin. In this first model, we use a general specification of TIME using index coding, and call the model ``M0i''.

Before we fit model M0i, we select the necessary columns from the data, and specify our priors. In the code of Tutorial 2a, model M0i is specified as follows.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{model\_M0i }\OtherTok{\textless{}{-}}                    
   \FunctionTok{brm}\NormalTok{(}\AttributeTok{data =}\NormalTok{ data\_M0i,}
       \AttributeTok{family =} \FunctionTok{bernoulli}\NormalTok{(}\AttributeTok{link=}\StringTok{"cloglog"}\NormalTok{),}
       \AttributeTok{formula =}\NormalTok{ event }\SpecialCharTok{\textasciitilde{}} \DecValTok{0} \SpecialCharTok{+}\NormalTok{ timebin }\SpecialCharTok{+}\NormalTok{ (}\DecValTok{0} \SpecialCharTok{+}\NormalTok{ timebin }\SpecialCharTok{|}\NormalTok{ pid),}
       \AttributeTok{prior =}\NormalTok{ priors\_M0i,}
       \AttributeTok{chains =} \DecValTok{4}\NormalTok{, }\AttributeTok{cores =} \DecValTok{4}\NormalTok{, }
       \AttributeTok{iter =} \DecValTok{3000}\NormalTok{, }\AttributeTok{warmup =} \DecValTok{1000}\NormalTok{,}
       \AttributeTok{control =} \FunctionTok{list}\NormalTok{(}\AttributeTok{adapt\_delta =} \FloatTok{0.999}\NormalTok{, }
                      \AttributeTok{step\_size =} \FloatTok{0.04}\NormalTok{, }
                      \AttributeTok{max\_treedepth =} \DecValTok{12}\NormalTok{),}
       \AttributeTok{seed =} \DecValTok{12}\NormalTok{, }\AttributeTok{init =} \StringTok{"0"}\NormalTok{,}
       \AttributeTok{file =} \StringTok{"Tutorial\_2\_Bayesian/models/model\_M0i"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

After selecting the bernoulli family and the cloglog link, the model formula is specified. The specification ``0 + \ldots{}'' removes the default intercept in brm(). The fixed effects include an intercept for each level of timebin. Each of these intercepts is allowed to vary across individuals (variable pid). Estimating model M0i took about 30 minutes on a MacBook Pro (Sonoma 14.6.1 OS, 18GB Memory, M3 Pro Chip).

\subsubsection{4.3.4 Model M1i: Adding the effects of prime-target congruency}\label{model-m1i-adding-the-effects-of-prime-target-congruency}

Previous research has shown that psychological effects typically change over time (Panis, 2020; Panis, Moran, et al., 2020; Panis \& Schmidt, 2022; Panis et al., 2017; Panis \& Wagemans, 2009). In the next model, we use index coding for both TIME (variable ``timebin'') and the categorical predictor prime-target-congruency (variable ``prime''), so that we get 30 grand intercepts, one for each combination of timebin level and prime level. Here is the model formula of this model that we call ``M1i''.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{event }\SpecialCharTok{\textasciitilde{}} \DecValTok{0} \SpecialCharTok{+}\NormalTok{ timebin}\SpecialCharTok{:}\NormalTok{prime }\SpecialCharTok{+}\NormalTok{ (}\DecValTok{0} \SpecialCharTok{+}\NormalTok{ timebin}\SpecialCharTok{:}\NormalTok{prime }\SpecialCharTok{|}\NormalTok{ pid)}
\end{Highlighting}
\end{Shaded}

Estimating model M1i took about 124 minutes.

\subsubsection{4.3.4 Model M1d: A more parsimonious model}\label{model-m1d-a-more-parsimonious-model}

When the shape of the hazard function is rather smooth, as it is for behavioral RT data, one can fit a more parsimonious model by treating TIME as a continuous variable, and use a polynomial specification of the effect of TIME. Thus, if we want to make assumptions about (1) how hazard changes over TIME in the reference condition (blank prime), and (2) how the effect of congruent and incongruent primes change over TIME (relax the proportionality assumption), then we can switch to a dummy coding approach for prime-target congruency (variable ``condition'') and treat TIME as a continuous variable (variable ``period\_9'').

For example, we may assume that hazard can change in a linear + quadratic fashion over time for a blank prime, and that the effects of congruent and incongruent primes relative to blank change in a linear + quadratic fashion, and fit the model called ``M1d''. Here is its model formula.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{event }\SpecialCharTok{\textasciitilde{}} \DecValTok{0} \SpecialCharTok{+}\NormalTok{ Intercept }\SpecialCharTok{+}\NormalTok{ condition}\SpecialCharTok{*}\NormalTok{period\_9 }\SpecialCharTok{+}\NormalTok{ condition}\SpecialCharTok{*}\NormalTok{period\_9\_sq }\SpecialCharTok{+} 
\NormalTok{                   (}\DecValTok{1} \SpecialCharTok{+}\NormalTok{ condition}\SpecialCharTok{*}\NormalTok{period\_9 }\SpecialCharTok{+}\NormalTok{ condition}\SpecialCharTok{*}\NormalTok{period\_9\_sq }\SpecialCharTok{|}\NormalTok{ pid)}
\end{Highlighting}
\end{Shaded}

The specification ``0 + Intercept + \ldots{}'' removes the default intercept in brm() and adds an explicit Intercept for which we can set the prior ourselves. Note that duplicate terms in the model formula (e.g., condition) are ignored.
Because TIME is centered on bin 9, the Intercept represents the estimated cloglog-hazard in bin 9 for the blank prime condition.

\subsubsection{4.3.5 Compare the models.}\label{compare-the-models.}

We can compare the three models using the Widely Applicable Information Criterion (WAIC) and Leave-One-Out (LOO) cross-validation, and look at model weights for both criteria (Kurz, 2023a; McElreath, 2018).

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{model\_weights}\NormalTok{(model\_M0i, model\_M1i, model\_M1d, }\AttributeTok{weights =} \StringTok{"loo"}\NormalTok{) }
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##     model_M0i     model_M1i     model_M1d 
## 2.462363e-183  1.000000e+00  5.739791e-17
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{model\_weights}\NormalTok{(model\_M0i, model\_M1i, model\_M1d, }\AttributeTok{weights =} \StringTok{"waic"}\NormalTok{) }
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##     model_M0i     model_M1i     model_M1d 
## 8.545948e-184  1.000000e+00  2.055767e-17
\end{verbatim}

Clearly, both the loo and waic weighting schemes assign a weight of 1 to model M1i, and a weight of 0 to the other two simpler models.

\subsubsection{4.3.6 Interpreting model M1i}\label{interpreting-model-m1i}

To make inferences from the parameter estimates in model M1i, we first plot the densities of the draws from the posterior distributions of its population-level parameters in Figure 5, together with point (median) and interval estimates (80\% and 95\% credible intervals).
